---
title: "PigWeb: behavioral indicators"
author: "Liza Moscoviche & Anja Eggert"
date: "`r Sys.Date()`" 
editor: visual
code-fold: false
toc: true
format: html
self-contained: true
execute:
  eval: true
  warning: false
  message: false
  echo: true
# project:
#   input-dir: "./PigWeb-behavior"
---

# Libraries

```{r, libraries}
library(kableExtra)
library(tidyverse)
library(rstatix)
```

# Data

## Read data

```{r}
dat <- read_csv("./data/PigWeb-behavior.csv",
                show_col_types = FALSE)
```

## Process data

```{r}
dat <- dat |> 
  mutate(across(where(is.character), as.factor)) |> 
  mutate(date           = dmy(date),
         time           = hms(time),
         institute      = fct_relevel(institute, "AU", "INRAE", "IRTA", "WUR"),
         institute      = relevel(institute, ref = "AU"),
         sample         = relevel(sample,    ref = "pre"),
         context        = relevel(context,   ref = "negative"),
         scans_inactive.prop = scans_inactive/scans_activity_coded,
         ear_forward.prop    = ear_forward/scans_ear_coded,
         ear_backward.prop   = ear_backward/scans_ear_coded)
```

# Descriptive statistics

## Locomotor play across institutes and context

```{r}
dat |> 
  group_by(institute, context) |> 
  rstatix::get_summary_stats(locomotor_play,
                             show = c("mean","sd")) |>
  ungroup() |> 
  arrange(variable, institute) |> 
  mutate(across(c(mean, sd), ~ round(., 2))) |> 
  kable(caption = "Table 3: Behavioral indicators") |>
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 10) |> 
  scroll_box()
```

## Escape events across institutes and context

```{r}
dat |> 
  group_by(institute, context) |> 
  rstatix::get_summary_stats(escape_events,
                             show = c("mean","sd")) |>
  ungroup() |> 
  arrange(variable, institute) |> 
  mutate(across(c(mean, sd), ~ round(., 2))) |> 
  kable(caption = "Table 3: Behavioral indicators") |>
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 10) |> 
  scroll_box()
```

## Proportion scans inactive across institutes and context

```{r}
dat |> 
  group_by(institute, context) |> 
  rstatix::get_summary_stats(scans_inactive.prop,
                             show = c("mean","sd")) |>
  ungroup() |> 
  arrange(variable, institute) |> 
  mutate(across(c(mean, sd), ~ round(., 2))) |> 
  kable(caption = "Table 3: Behavioral indicators") |>
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 10) |> 
  scroll_box()
```

## Proportion scans ears forward across institutes and context

```{r}
dat |> 
  group_by(institute, context) |> 
  rstatix::get_summary_stats(ear_forward.prop,
                             show = c("mean","sd")) |>
  ungroup() |> 
  arrange(variable, institute) |> 
  mutate(across(c(mean, sd), ~ round(., 2))) |> 
  kable(caption = "Table 3: Behavioral indicators") |>
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 10) |> 
  scroll_box()
```

## Proportion scans ears backward across institutes and context

```{r}
dat |> 
  group_by(institute, context) |> 
  rstatix::get_summary_stats(ear_backward.prop,
                             show = c("mean","sd")) |>
  ungroup() |> 
  arrange(variable, institute) |> 
  mutate(across(c(mean, sd), ~ round(., 2))) |> 
  kable(caption = "Table 3: Behavioral indicators") |>
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 10) |> 
  scroll_box()
```

# Citing R

"All analyses were performed using R Statistical Software (version 4.4.2; R Core Team 2024)".

Reference: R Core Team (2024). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.

```{r, R-Zitieren}
citation()
version$version.string
```

```{r, Pakete-Zitieren}
citation("kableExtra")
citation("tidyverse")
citation("rstatix")
```

# Session Info

```{r, R-Session}
sessionInfo()
```
