---
title: "PigWeb: behavior-ears"
author: "Liza Moscoviche & Anja Eggert"
date: "`r Sys.Date()`" 
editor: visual
code-fold: false
toc: true
format: html
self-contained: true
execute:
  eval: true
  warning: false
  message: false
  echo: true
# project:
#   input-dir: "./PigWeb-behavior"
---

# Libraries

```{r, libraries}
library(kableExtra)
library(tidyverse)
library(rstatix)
library(glmmTMB)
library(car)
library(DHARMa)
library(patchwork)
```

# Scripts, etc.

```{r}
glmmTMBcontr = glmmTMBControl(optimizer = optim, 
                              optArgs   = list(method="BFGS"))
```

```{r}
my_theme = theme_classic() +
  theme(text = element_text(family = "sans"),
        plot.title   = element_text(face = "plain", size = 16),
        axis.title   = element_text(face = "bold", size = 16),
        axis.text    = element_text(size = 14, angle = 0, vjust = 0.5),
        legend.title = element_text(face = "bold", size = 16),
        legend.text  = element_text(face = "plain", size = 16),
        strip.text   = element_text(size = 16),
        legend.position = "top",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

```{r}
set.seed(1989)
```

# Data

## Read data

```{r}
dat <- read_csv("./data/PigWeb-behavior.csv",
                show_col_types = FALSE)
```

## Process data

```{r}
dat <- dat |> 
  mutate(across(where(is.character), as.factor)) |> 
  mutate(date           = dmy(date),
         date.f         = as.factor(date),
         time           = hms(time),
         institute      = fct_relevel(institute, "AU", "INRAE", "IRTA", "WUR"),
         institute      = relevel(institute, ref = "AU"),
         sample         = relevel(sample,    ref = "pre"),
         context        = relevel(context,   ref = "negative"),
         time_collect   = hour(time) + minute(time)/60,
         time_collect.z = as.vector(scale(time_collect)),
         space_allowance       = pen_space_m2/(group_weight_kg^0.66),
         space_allowance.log.z = as.vector(scale(log(space_allowance)))) |> 
  mutate(prop_ear_forward = (ear_forward/scans_ear_coded),
         prop_ear_backward = (ear_backward/scans_ear_coded),
         # small transformation to prevent zero's in the data
         prop_ear_forward.trans  = (prop_ear_forward*(nrow(dat)-1)+0.5)/nrow(dat),
         prop_ear_backward.trans = (prop_ear_backward*(nrow(dat)-1)+0.5)/nrow(dat)) 
```

# Descriptive statistics across context

Pooled for all institutes:

```{r}
dat |> 
  group_by(context) |> 
  rstatix::get_summary_stats(prop_ear_forward, prop_ear_backward,
                             show = c("mean","sd")) |>
  ungroup() |> 
  arrange(variable) |> 
  mutate(across(c(mean, sd), ~ round(., 2))) |> 
  kable(caption = "Text for Fig. 6") |>
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 12) |> 
  scroll_box()
```

# Both ears forward

## Binomial model: no interactions

Fitting model on proportion of ears forward of all scans (transformed values with no zero's). Model without interactions of fixed effects and random intercepts but constant slope.

Full model:

```{r}
mod.ears.forw <- glmmTMB(prop_ear_forward.trans ~
                           context +
                           institute +
                           space_allowance.log.z + 
                           offset(log(scans_ear_coded)) +
                           (1|date) + 
                           (1|pig_group_unique) + 
                           (1|pig_id_unique),
                         data      = dat,
                         family    = beta_family(link = "logit"), 
                         control   = glmmTMBcontr)
```

Reduced model, without `context`:

```{r}
mod.ears.forw.red <- glmmTMB(prop_ear_forward.trans ~
                               institute +
                               space_allowance.log.z + 
                               offset(log(scans_ear_coded)) +
                               (1|date) + 
                               (1|pig_group_unique) + 
                               (1|pig_id_unique),
                             data      = dat,
                             family    = beta_family(link="logit"), 
                             control   = glmmTMBcontr)
```

Comparison of reduced and full model:

```{r}
anova(mod.ears.forw.red, mod.ears.forw, test = "Chisq")
```

### Final model

Model summary, focus on estimates:

```{r}
round(summary(mod.ears.forw)$coefficients$cond, 2)
```

Fixed effects

```{r}
round(drop1(mod.ears.forw,
            test = "Chisq"), 3)

car::Anova(mod.ears.forw,
           test.statistic = "Chisq",
           type = 2) # as no interaction, otherwise set type = 3
```

### Model diagnostics

Collinearity:

```{r}
xx=lm(prop_ear_forward.trans ~ 
        context + 
        institute + 
        space_allowance.log.z, 
      data = dat)
car::vif(xx) 
```

Zero inflation:

```{r}
DHARMa::testZeroInflation(mod.ears.forw)
```

# Both ears backward

## Binomial model: no interactions

Fitting model on proportion of ears backward of all scans (transformed values with no zero's). Model without interactions of fixed effects and random intercepts but constant slope.

Full model:

```{r}
mod.ears.back <- glmmTMB(prop_ear_backward.trans ~
                           context +
                           institute +
                           space_allowance.log.z + 
                           offset(log(scans_ear_coded)) +
                           (1|date) + 
                           (1|pig_group_unique) + 
                           (1|pig_id_unique),
                         data      = dat,
                         family    = beta_family(link = "logit"), 
                         control   = glmmTMBcontr)
```

Reduced model, without `context`:

```{r}
mod.ears.back.red <- glmmTMB(prop_ear_backward.trans ~
                               institute +
                               space_allowance.log.z + 
                               offset(log(scans_ear_coded)) +
                               (1|date) + 
                               (1|pig_group_unique) + 
                               (1|pig_id_unique),
                             data      = dat,
                             family    = beta_family(link = "logit"), 
                             control   = glmmTMBcontr)
```

Comparison of reduced and full model:

```{r}
anova(mod.ears.back.red, mod.ears.back, test = "Chisq")
```

### Final model

Model summary

```{r}
round(summary(mod.ears.back)$coefficients$cond, 2)
```

Fixed effects

```{r}
round(drop1(mod.ears.back,
            test = "Chisq"), 3)

car::Anova(mod.ears.back,
           test.statistic = "Chisq",
           type = 2) # as no interaction, otherwise set type = 3
```

### Model diagnostics

Collinearity:

```{r}
xx <- lm(prop_ear_backward.trans ~ 
        context + 
        institute + 
        space_allowance.log.z, 
      data = dat)
car::vif(xx) 
```

Zero inflation:

```{r}
DHARMa::testZeroInflation(mod.ears.back)
```

# Figures

## Both ears forward

There are 35 NA's in the data:

```{r}
summary(dat$prop_ear_forward)
```

```{r}
plot.forw <- dat |> 
  mutate(context_jit = jitter(as.numeric(context), 0.3)) |> 
  
  ggplot(aes(y = prop_ear_forward)) + 
  geom_boxplot(aes(x = context),
               col = "black", fill = NA, 
               outlier.shape = NA, width=0.5,
               position = position_dodge(0.5)) +
  geom_line(aes(x = context_jit, group = pig_id_unique), 
            col = "grey80", linewidth = 0.2) + 
  geom_point(aes(x = context_jit, 
                 col = institute, 
                 shape = institute), 
             size = 3, alpha = 0.7) + 
  scale_colour_manual(values = c("#481568FF", "#32648EFF", "#94D840FF", "#FDE725FF")) +
  scale_shape_manual(values = c(17, 15, 16, 18)) +
  scale_y_continuous(limits=c(-0.1, 1.1), 
                     breaks = seq(0,1,0.2)) +
  labs(x   = "Context",
       y   = "Ears forward (proportions)",
       # Matching title ensures a single legend
       col = "Institutes",
       shape = "Institutes") +
  my_theme
```

```{r}
plot.forw
```

```{r, eval = FALSE}
png("./plots/ears-forward-prop-context.png",
    width = 150, height = 180, units = "mm",
    pointsize = 10, res = 600)

plot.forw

dev.off()
```

## Both ears backward

There are 35 NA's in the data:

```{r}
summary(dat$prop_ear_backward)
```

```{r}
plot.back <- dat |> 
  mutate(context_jit = jitter(as.numeric(context), 0.3)) |> 
  
  ggplot(aes(y = prop_ear_backward)) + 
  geom_boxplot(aes(x = context),
               col = "black", fill = NA, 
               outlier.shape = NA, width=0.5,
               position = position_dodge(0.5)) +
  geom_line(aes(x = context_jit, group = pig_id_unique), 
            col = "grey80", linewidth = 0.2) + 
  geom_point(aes(x = context_jit, 
                 col = institute, 
                 shape = institute), 
             size = 3, alpha = 0.7) + 
  scale_colour_manual(values = c("#481568FF", "#32648EFF", "#94D840FF", "#FDE725FF")) +
  scale_shape_manual(values = c(17, 15, 16, 18)) +
  scale_y_continuous(limits=c(-0.1, 1.1), 
                     breaks = seq(0,1,0.2)) +
  labs(x   = "Context",
       y   = "Ears backward (proportions)",
       # Matching title ensures a single legend
       col = "Institutes",
       shape = "Institutes") +
  my_theme
```

```{r}
plot.back
```

```{r, eval = FALSE}
png("./plots/ears-backward-prop-context.png",
    width = 150, height = 180, units = "mm",
    pointsize = 10, res = 600)

plot.back

dev.off()
```

## Figure combined: (A) ears forward and (B) ears backward

```{r}
combined <- plot.forw + plot.back & 
  theme(legend.position = "top")

combined + 
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = 'A') & 
  theme(plot.tag = element_text(size = 16),
        plot.tag.position = c(.15, .88))
```

```{r, eval = FALSE}
png("./plots/ears-forw-backw-prop-context.png",
    width = 300, height = 180, units = "mm",
    pointsize = 10, res = 600)

combined + 
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = 'A') & 
  theme(plot.tag = element_text(size = 16),
        plot.tag.position = c(.15, .88))

dev.off()
```

# Citing R

"All analyses were performed using R Statistical Software (version 4.4.2; R Core Team 2024)".

Reference: R Core Team (2024). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.

```{r, R-Zitieren}
citation()
version$version.string
```

```{r, Pakete-Zitieren}
citation("kableExtra")
citation("tidyverse")
citation("rstatix")
citation("glmmTMB")
citation("car")
citation("DHARMa")
citation("patchwork")
```

# Session Info

```{r, R-Session}
sessionInfo()
```
